---
title: "Case Study 2"
subtitle: "AKSTA Statistical Computing"
author:
  - Carla Salazar
  - Veneta Grigorova
  - Juraj Simkovic
date: "2025-26-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
*The  .Rmd* **and** *.html (or .pdf) should be uploaded in TUWEL by the deadline. Refrain from using explanatory comments in the R code chunks but write them as text instead. Points will be deducted if the submitted file is not in a decent form.*

**DISCLAIMER**: In case students did not contribute equally, include a disclaimer stating what each student's contribution was.


The CIA World Factbook provides intelligence on various aspects of 266 world entities, including history, people, government, economy, energy, geography, environment, communications, transportation, military, terrorism, and transnational issues. This case study involves analyzing world data from 2020, focusing on:

- **Education Expenditure (% of GDP)**
- **Youth Unemployment Rate (15-24 years)**
- **Net Migration Rate** (difference between the number of people entering and leaving a country per 1,000 persons)

The data was sourced from the [CIA World Factbook Archives](https://www.cia.gov/the-world-factbook/about/archives/). You are required to use `dplyr` for data manipulation, while any package can be used for importing data.

# Tasks:

## a. Data Import and Cleaning

Load the following datasets from TUWEL and ensure that missing values are handled correctly and column names are clear. Each dataset should ultimately contain only two columns: **country** and the respective variable. Note that some data sets also contain information on the year when the value was last updated.

  
  * `rawdata_369.txt` which contains the (estimated)  public expenditure on education as a percent of GDP. *Pay attention! The delimiter is 2 or more white spaces (one space would not work as it would separate country names which contain a space); you have to skip the first two lines*. 
  
  * `rawdata_373.csv` which contains the (estimated) youth unemployment rate (15-24) per country
  
  * `rawdata_347.txt`  which contains (estimated) net migration rate per country.
  
```{r}
library(dplyr)
library(readr)
library(stringr)
library(countrycode)
library(readxl)
library(readr)
library(stringi)
library(dplyr)
```

```{r}
# education expenditure
edu_raw <- readLines("data/rawdata_369.txt")[-(1:2)]  # skip first 2 lines
edu_df <- read.table(text = edu_raw, sep = "", fill = TRUE, strip.white = TRUE, 
                     stringsAsFactors = FALSE, header = FALSE)

head(edu_df)

# parse education data
edu_parsed <- str_match(edu_raw, "^(\\d+)\\s+(.+?)\\s+(\\d+\\.\\d+)\\s+(\\d{4})$")

# convert to clean data frame
edu_df_clean <- data.frame(
  country = edu_parsed[, 3],
  expenditure = as.numeric(edu_parsed[, 4]),
  stringsAsFactors = FALSE
)

edu_df_clean
```

```{r}
# net migration rate
mig_raw <- readLines("data/rawdata_347.txt")[-(1:2)]

# use regex to extract rank, country, migration rate, and year
mig_parsed <- str_match(mig_raw, "^(\\d+)\\s+(.+?)\\s+(-?\\d+\\.\\d+)\\s+(\\d{4})")

# build clean df
mig_df_clean <- data.frame(
  country = mig_parsed[, 3],
  net_migration = as.numeric(mig_parsed[, 4]),
  stringsAsFactors = FALSE
)

mig_df_clean
```

```{r}
# rename columns for consistency
unemp_df_clean <- read_csv("data/rawdata_373.csv") %>%
  select(country = country_name,
         youth_unemployment = youth_unempl_rate)

unemp_df_clean
```

## b. Merging Raw Data

Merge the datasets using `dplyr` on a unique key and retain the union of all observations.

```{r}
library(dplyr)

# full join education and unemployment
df_merge1 <- full_join(edu_df_clean, unemp_df_clean, by = "country")

# join net migration
df_all <- full_join(df_merge1, mig_df_clean, by = "country")

dim(df_all)
head(df_all)
```

- What key are you using for merging?

Key used for merging: country (it is common in all 3 datasets)

- Return the dimensions of the merged dataset.

Dimensions of merged dataset: 227 rows × 4 columns

## c. Enriching Data with Income Classification

Obtain country income classification (low, lower-middle, upper-middle, high) from the [World Bank](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519) and merge it with the dataset.

- Identify common variables between datasets. Can they be used for merging? Why or why not?

The datasets all have country names in common, but the problem is that these names aren't always written the same way. For example, one file might say "Venezuela" while another says "Venezuela, RB", or "Korea, South" vs "Korea, Rep." Because of these differences in spelling and formatting, we can’t reliably merge the datasets just using country names unless we clean them up first

- Since ISO codes are standardized, download and use the [CIA country data codes](https://www.cia.gov/the-world-factbook/references/country-data-codes/) for merging. Make sure you are not losing any of the countries in your original data set when merging.

```{r}
# patch country names before standardization
df_all <- df_all %>%
  mutate(
    country = case_when(
      country == "Saint Martin" ~ "St. Martin (French part)",
      country == "Virgin Islands" ~ "British Virgin Islands",
      TRUE ~ country
    )
  )

# use countrycode to standardize names and get ISO-3 codes
df_all_iso <- df_all %>%
  mutate(
    country_standard = countrycode(country, origin = "country.name", destination = "country.name"),
    iso3c = countrycode(country, origin = "country.name", destination = "iso3c")
  ) %>%
  mutate(
    iso3c = case_when(
      country == "Kosovo" ~ "XKX",
      country == "Venezuela" ~ "VEN",
      country == "St. Martin (French part)" ~ "MAF",
      country == "British Virgin Islands" ~ "VGB",
      TRUE ~ iso3c
    )
  )

# load income classification (World Bank)
income_df <- read_excel("data/CLASS.xlsx", sheet = 1) %>%
  select(country = Economy,
         iso_code = Code,
         income = `Income group`) %>%
  mutate(iso_code = str_trim(iso_code))

# join income classification using ISO-3 codes
df_all_income <- df_all_iso %>%
  left_join(income_df, by = c("iso3c" = "iso_code"))

# mark unmatched income classifications
df_all_income <- df_all_income %>%
  mutate(income = ifelse(is.na(income), "Not classified", income))

df_all_income

df_all_income_clean <- df_all_income %>%
  select(
    country = country_standard,     # clean country name
    iso = iso3c,                    # standard ISO code
    income,                         # World Bank income classification
    expenditure,
    youth_unemployment,
    net_migration
  )

df_all_income_clean
glimpse(df_all_income_clean)
```

## d.  Adding Geographical Information

Introduce continent and subcontinent (or region) data for each country.

- Find and download an appropriate online resource.
- Merge this information into the dataset, naming the final dataset `df_vars`. Make sure you are not losing any of the countries in your original data set when merging.

```{r}
# load continent mapping
continent_data <- read.csv("data/countryContinent.csv")
names(continent_data)

# load continent info and rename relevant columns
continent_data <- read.csv("data/countryContinent.csv") %>%
  select(
    iso = code_3,
    continent,
    region = sub_region
  ) %>%
  mutate(iso = str_trim(iso))

# merge with cleaned dataset using ISO-3
df_vars <- df_all_income_clean %>%
  left_join(continent_data, by = "iso")

df_vars

# look for unmatched
df_vars %>%
  filter(is.na(continent)) %>%
  select(country, iso) %>%
  distinct()

# patch Kosovo region and continent
df_vars <- df_vars %>%
  mutate(
    continent = ifelse(iso == "XKX", "Europe", continent),
    region = ifelse(iso == "XKX", "Southern Europe", region)
  )
```


## e. Data Tidiness and Summary Statistics

- Evaluate the tidiness of `df_vars`
(observational units, variables, fixed vs. measured variables).
Make adjustments to tidy the data, if necessary.

```{r}
glimpse(df_vars)
```

Each row in the dataset stands for one country, and every column contains a specific variable like expenditure, unemployment, migration, or income. The data already looks clean and organized, so we don’t need to reshape or tidy it further.

- Create a frequency table for the income status variable and briefly interpret the results.

```{r}
# absolute frequency
table(df_vars$income)

# relative frequency
prop.table(table(df_vars$income))
```
The dataset has 84 high income countries, 54 upper middle income, 52 lower middle income, 26 low income, and 10 countries that are not classified.
High income countries make up about 37% of the data, followed by upper middle (24%), lower middle (23%), and low income (11%). Only around 4% of countries are not classified.
This shows the data is somewhat balanced, but there's a clear tilt toward middle and high income countries.

- Analyze the distribution of income status across continents by computing absolute and relative frequencies. Comment on the findings.

```{r}
# absolute counts
table(df_vars$continent, df_vars$income)

# relative frequencies (row-wise proportions)
prop.table(table(df_vars$continent, df_vars$income), margin = 1)
```
In **Africa** (55 countries), 22 are low income (40%) and 23 are lower middle income (42%). This means over 80% of African countries fall into the lower income brackets. Only 1 country is high income (2%). This reflects ongoing economic development challenges across much of the continent.

In the **Americas** (49 countries), 21 are high income (43%) and 19 are upper middle income (39%). 4 are lower middle income (8%) and 5 are not classified (10%). The region is clearly skewed toward wealthier countries, but a few countries still fall into the middle and unclassified groups. This mix reflects strong economies like the US and Canada, alongside developing nations in Central and South America.

**Asia** (52 countries) is the most economically diverse. It has 15 high income (29%), 14 upper middle (27%), 19 lower middle (37%), and 4 low income countries (8%). The range covers some of the world’s richest countries (like Japan and South Korea) and some of the poorest (like Afghanistan and Yemen), showing Asia’s economic contrasts.

**Europe** (49 countries) is heavily dominated by high income countries — 38 of them (78%). The rest are mostly upper middle income (18%), with just 2 not classified. There are no low or lower middle income countries. This confirms Europe’s position as one of the wealthiest regions in the world.

**Oceania** (21 countries) has 9 high income (43%) and 6 lower middle income (29%), with 4 upper middle (19%) and 2 unclassified (10%). The split here shows the contrast between rich countries like Australia and New Zealand, and smaller island nations with lower economic status.

The data reflects global economic inequalities. Wealth is concentrated in Europe and parts of Oceania, while Africa remains largely low income. Asia and the Americas are more balanced, containing both very rich and developing countries. These patterns align with global development trends and regional histories.

- Using the distribution of income status across continents, identify which countries are the only ones in their income group across the continent. Discuss briefly.

```{r}
# find unique income group within each continent
df_vars %>%
  filter(income != "Not classified") %>%
  group_by(continent, income) %>%
  filter(n() == 1) %>%
  select(country, continent, income) %>%
  arrange(continent, income)
```

```{r}
# find rare income groups within each continent
df_vars %>%
  filter(income != "Not classified") %>%
  group_by(continent, income) %>%
  filter(n() <= 5) %>%
  select(country, continent, income) %>%
  arrange(continent, income)
```

These countries differ from the economic pattern of their continent for several reasons:

- **Seychelles** has a small population and benefits from high-end tourism and stable governance, which helps it achieve high income despite being in Africa where most countries are lower income.

- In the **Americas**, countries like *Bolivia*, *Honduras*, *Nicaragua*, and *Haiti* have weaker infrastructure, lower education levels, and histories of political instability or colonial exploitation. These factors contribute to them lagging behind wealthier neighbors.

- **Asia’s** low income countries — *Afghanistan*, *Syria*, *Yemen*, and *North Korea* — are heavily affected by war, sanctions, or authoritarian regimes. These issues limit foreign investment, trade, and development.

- **Fiji**, although a small island nation, has a relatively strong economy for the Pacific thanks to tourism, remittances, and regional political influence. It stands apart from nearby countries that rely more heavily on subsistence agriculture or aid.

In general, these cases show how **politics, geography, conflict, and resource access** all play major roles in shaping a country’s economic status — not just its location on the map.

## f. Further Summary Statistics and Insights

- Create a table of average (mean and median) values for expenditure, youth unemployment rate and net migration rate separated into income status. Make sure that in the output, the ordering of the income classes is proper (i.e., L, LM, UM, H or the other way around). Briefly comment the results and any differences between the mean and median. 

- Look at the standard deviation and the interquartile range of the variables per income status instead of the location statistics above. Do you gain additional insights? Briefly comment the results. 


- Extend the analysis of the statistics median and IQR  to **each income status and continent combination**. Play around with displaying the resulting table. Use `pivot_longer()` and/or 
`pivot_wider()` to generate different outputs. 
Discuss the results as well as the readability of the different tables.
 
- Identify countries performing well in terms of both **youth unemployment** and **net migration rate** (top 25% in net migration and bottom 25% in youth unemployment within their continent).

## g. Conditional Probabilities

Estimate the following based on the observed frequencies in the data:

- What is the (posterior or conditional) probability that a European country belongs to the high income group? What is the prior probability that a country belongs to the high income group?

- Given a country has high youth unemployment (above %25), what is the probability that it also has negative net migration?


## h. Simpson’s Paradox Analysis

Investigate whether an overall trend in youth unemployment rate in the high and low income groups reverses when analyzed at the continent level. E.g., does the youth unemployment rate appear lower in low-income countries overall, but higher when controlling for continent? Explain the results and possible reasons behind this paradox.


## i. Data Export

Export the final tidy dataset from e. as a **CSV** with:`;` as a separator; `.` representing missing values; no row names included.  Upload the `.csv` to TUWEL, together with the submission.

